Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
last improvements

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
  Tabular data proves to be difficult for neural networks, where they are still outperformed by other model classes.
  Following recent research, regularisation can improve their performance. Incorporating the EAGGA framework designed to
  enforce explainability, we improve interpretability and achieve performance comparable to that of XGBoost on the
  metrics AUC, number of features, number of pairwise interactions, and number of non-monotone feature effects.
  For this, we propose a sparse, regularised network architecture specifically suited for EAGGA's constraints on
  interaction and monotonicity.
\end{abstract}

@@

\section{Introduction}
Regularisation has been extensively shown to improve performance from linear models to neural networks.
One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
multicollinear data in linear models \citep[p. 55]{ridge}.
Practicioners training neural networks regularly use regularisation methods such as dropout to reduce
co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing} or early stopping to avoid overfitting on the training data \citep[p. 778f]{early_stopping}.
\\
Some regularisation can even increase interpretability, such as LASSO, or L1-regularisation for linear models, which reduces the number of
features used to fit the model by setting some coefficients to 0 \citep[p. 267]{lasso}.
\\
Neural networks as a model class have achieved impressive performance on a multitude of problems, yet still struggle with the most common format --
tabular data \citep[p. 7499]{Borisov_2024}.
Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[p. 8]{NEURIPS2021_c902b497}.
\citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
In this report, we explore if the same framework can be extended to neural networks, a model class that in its fully-connected feedforwad form is
inherently non-interpretable, to tackle both explainability and achieve performance on tabular data that is on par with XGBoost.

@@

% Why Interpretability is desirable?
Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3f]{Zach2019InterpretabilityOD} gives a range of examples,
amongst which are \textit{gaining trust}, e.g. when doctors rely on medical diagnosis predictions, avoiding \textit{subconcious biases} by
making sure loan approvals are non-discriminatory, or \textit{regulatory} reasons, most notably the EU's ``Right to Explanation'' warranted by the
GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[pp. 727f]{survey_NN_interpretability}.
It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimizing a model with respect to loss and then judging its
performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[pp. 727f]{survey_NN_interpretability} or for
\textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
made accessible to humans to be used reliably \citep[p. 1]{review_NN_interpretability}.

@@

% Taxonomy
\citet[pp. 728f]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs).
\\
\textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
training process to increase model interpretability.
\\
\textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.
\\
\textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
explaining model behaviour for sets of samples grouped by some criterion, to \textit{global} methods, which explain the network as a whole.

% Evaluation
Given its unclear definition, evaluating interpretability can be challenging. \citet[pp. 4-6]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
possible evaluation methods based on their rigorousness.
\\
\textbf{Application-grounded evaluation} evaluates interpretations directly with respect to the task, by having human experts evaluate the outcome and
is therefore the most expensive and time-consuming of the three approaches.
\\
\textbf{Human-grounded metrics} is similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
so that a layperson can do it. Human-grounded approaches are especially suitable if it's sufficient to validate the general concepts of a task. A typical evaluation
set-up in this category is binary forced choice, where the human evaluator chooses, which of two generated explanations he prefers.
\\
\textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assess explanatory quality according to some
formally defined proxy for interpretability and is particularly useful in ranking different models if their model-class is already identified
(e.g. via human-grounded evaluation) to be interpretable. The main challenge for functionally-grounded evaluation is finding a good proxy.

\subsection{Hyperparameter Optimization (HPO)}
In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are
fixed before training. Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration a chosen loss function with $L$.
The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
For this, there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}

@@

% model free
% grid + random search
The most basic model-free strategies for blackbox optimization are grid and random search.
In the latter, the user defines a range of interest for each HP and then evaluating points along a grid within the cartesian product of those.
This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions
about the model makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance,
as for a given budget $B$, each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$
different values for $N$ HPs. \citep[pp. 7f]{feurer_hyperparameter_2019}

% evolutionary algorithms
Another popular class of model-free optimizers are evolutionary algorithms (EAs), which are conceptually simple and can handle even complex parameter spaces,
given appropriate implementation of operators.
EAs are iteratively evolving a population of individuals (an individual is simply a hyperparameter configuration) of size $\mu$, where in each iteration (generation),
$\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-14]{genetic_algos}.
\\
\textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
performance metric the model resulting from the individual's HP configuration is evaluated on.
\\
\textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
index $k$ on are swapped, yielding two new ``children'' individuals.
\\
\textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
on binary values.
\\
At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
\citep[p. 188]{simon2013evolutionary}

@@

Another popular EA approach is Differential Evolution. Its populaton is randomly initialized so the entire search space is covered.
In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
$\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})^T$ with
\begin{equation}
  \upsilon_{i,g+1}^{(j)} = \begin{dcases}
    \nu_{i,g+1}^{(j)} & \text{if } u\le\text{CR}\text{ or } j=R \\
    \lambda_{i,g}^{(j)} & \text{else}
  \end{dcases}
\end{equation}
for a random index $R\in\{1,2,...,D\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}

@@

The kernel function can be freely chosen, a common choices is the Gaussian (Eq. \ref{eq-kernel-gaussian}) kernel with

@@

A drawback of Gaussian Processes is its poor scalability in number of data points and number of hyperparameters, although there exist workarounds approximating the full GP

@@

\subsection{Neural Architecture Search}
A special case of HPO is Neural Architecture Search (NAS). Because of the flexibility of neural network architectures (number of hidden layers, strength of dropout,
each layer can have a different numer of neurons, each layer can have a different activation function, etc.), traditional HPO methods cannot efficiently explore the
entire space.
In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and
image domains \citep{baker2017designingneuralnetworkarchitectures},  \citep{10.5555/3504035.3504375},  \citep{10.1145/3071178.3071229},
where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spaces}.
These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For regular feedforwad neural networks,
this could be a linear layer of fixed size and with a specific activation function,
for CNNs this could be a specific convolutional operation or pooling function, for instance.
Instead of tuning every hyperparameter, the sequential placement of the cells is then optimzed during NAS, e.g. via random search or Bayesian
Optimization. \citep[p. 67]{elsken_neural_2019}
\\
Treating a neural network as a sequence of cells instead of one fixed entity further allows exploiting this structure with special sequential optimization techniques.
\citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} and \citet[pp. 2-4]{Zoph_2018_CVPR} propose a cell search method using RNNs and reinforcement learning.
The cells in their approach are not fixed, but have hyperparameters themselves, such as the filter size and the stride size of a convolutional layer.
These hyperparameters are then predicted sequentially, given all the already predicted hyperparameters of previous layers. The RNN is trained using reinforcement
learning, where the actions are the different values the network can predict for a given hyperparameter and the reward function is the RNN's performance on held-out
validation data.
\\
The second approach to NAS is the \textbf{one-shot model}, which trains a ``fabric'', which can be understood as a DAG comprising all architectures of a given
search space. Each path through the DAG is one specific network architecture, where the nodes represent a layer of neurons and the edges in-between are operations
on the layers' values. The entire DAG has one designated input and one output node.
It is then trained just like a regular NN, after which the optimal path, i.e. network architecture, is selected from the fabric.
This has the advantage of, albeit being slower than training a single model, being far more efficient and less expensive than training all architectures encoded
in the fabric. \citep[pp. 1-2, p. 8]{saxena2017convolutionalneuralfabrics}
% don't necessarily talk about co-adaptation problem

@@

\subsubsection{A-posteriori}
\label{sec-moo-post}
The disadvantage of a-priori methods is the missing knowledge of the interplay between a hyperparameter configuration and the trained model's
performance across objectives:
we can either restrict the search space to enforce some will, such as only using a maximum of 50\% of features, optimize for performance and use the
resulting model, or leave the search space unrestricted, adjust the loss function to incorporate all objectives and take its optimum.
In neither case do we know the impact our a-priori trade-off has on the final performance.
In practical applications, the impact on performance across all objectives is what matters.
This is the main advantage of a-posteriori methods: instead of implicitly defining a trade-off prior to training, we still evaluate multiple HP configurations
but keep the set of non-dominated solutions instead of just one that satisfies the trade-off criterion.
This is beneficial as it makes the relationship visible, a practicioner can now see, for instance, that a slight decrease in one objective might translate to
a significant improvement in another that would have been missed if the problem was optimized with a-priori methods.
\\
Aside from the usual baselines grid and random search there are multi-objetive Bayesian Optimization adaptations, mainly using one of two approaches:
\begin{enumerate}[label*=\arabic*.]
  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augmented Tchebycheff function as
        scalarization to ensure the Pareto front is explored sufficiently.
  \item Fitting one surrogate per objective, then either 
  \begin{enumerate}[label*=\arabic*.]
    \item using one acquisition function per surrogate to return a set of promising candidate query points or
    \item using one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises
          the expected improvement of hypervolume.
  \end{enumerate}
\end{enumerate}

>>>>

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
  Tabular data proves to be difficult for neural networks, where they are still outperformed by other model classes.
  Following recent research, regularisation can improve their performance. Incorporating the EAGGA framework designed to
  enforce explainability, we improve interpretability and achieve performance comparable to that of XGBoost on
  dominated hypervolume on the metrics AUC, number of features, number of pairwise interactions, and number of non-monotone feature effects.
  For this, we propose a sparse, regularised network architecture specifically suited for EAGGA's constraints on
  interaction and monotonicity.
\end{abstract}

@@

\section{Introduction}
Regularisation has been extensively shown to improve performance from linear models to neural networks (NNs).
One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
multicollinear data in linear models \citep[p. 55]{ridge}.
Practicioners training neural networks regularly use regularisation methods such as dropout to reduce
co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing} or early stopping to avoid overfitting on the training data \citep[p. 778f]{early_stopping}.
\\
Some regularisation can even increase interpretability, such as LASSO, or L1-regularisation for linear models, which reduces the number of
features used to fit the model by setting some coefficients to 0 \citep[p. 267]{lasso}.
\\
Neural networks as a model class have achieved impressive performance on a multitude of problems, yet still struggle with the most common format --
tabular data \citep[p. 7499]{Borisov_2024}.
Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[p. 8]{NEURIPS2021_c902b497}.
\citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
In this report, we explore if the same framework can be extended to neural networks, a model class that in its fully-connected feedforwad form is
inherently non-interpretable, to tackle both explainability and achieve performance on tabular data that is on par with XGBoost.

@@

% Why Interpretability is desirable?
Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3f]{Zach2019InterpretabilityOD} gives a range of examples,
amongst which are \textit{gaining trust}, e.g. when doctors rely on medical diagnosis predictions, avoiding \textit{subconcious biases} by
making sure loan approvals are non-discriminatory, or \textit{regulatory} reasons, most notably the EU's ``Right to Explanation'' warranted by the
GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[pp. 727f]{survey_NN_interpretability}.
It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimizing a model with respect to loss and then judging its
performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[pp. 727f]{survey_NN_interpretability} or for
\textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
made accessible to humans in order to be used reliably \citep[p. 1]{review_NN_interpretability}.

@@

% Taxonomy
\citet[pp. 728f]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks.
\\
\textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
training process to increase model interpretability.
\\
\textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.
\\
\textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
explaining model behaviour for sets of samples grouped by some criterion, to \textit{global} methods, which explain the network as a whole.

% Evaluation
Given its unclear definition, evaluating interpretability can be challenging. \citet[pp. 4-6]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
possible evaluation methods based on their rigorousness.
\\
\textbf{Application-grounded evaluation} evaluates interpretations directly with respect to the task, by having human experts evaluate the outcome and
is therefore the most expensive and time-consuming of the three approaches.
\\
\textbf{Human-grounded metrics} are similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
so that a layperson can do it. Human-grounded approaches are especially suitable if it's sufficient to validate the general concepts of a task. A typical evaluation
set-up in this category is \textit{binary forced choice}, where the human evaluator chooses, which of two generated explanations they prefer.
\\
\textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assesses explanatory quality according to some
formally defined proxy for interpretability and is particularly useful in ranking different models if their model-class is already identified
(e.g. via human-grounded evaluation) to be interpretable. The main challenge for functionally-grounded evaluation is finding a good proxy.

\subsection{Hyperparameter Optimization (HPO)}
In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are
fixed before training. Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration $\boldsymbol\lambda$ and a chosen loss function with $L$.
The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
For this, there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}

@@

% model free
% grid + random search
The most basic model-free strategies for blackbox optimization are grid and random search.
In the latter, the user defines a range of interest for each HP and then evaluates points along a grid within the cartesian product of those.
This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions
about the model make it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance,
as for a given budget $B$, each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$
different values for $N$ HPs. \citep[pp. 7f]{feurer_hyperparameter_2019}

% evolutionary algorithms
Another popular class of model-free optimizers are evolutionary algorithms (EAs), which are conceptually simple and can handle even complex parameter spaces,
given appropriate implementation of operators.
EAs are iteratively evolving a population of individuals (an individual is simply a hyperparameter configuration) of size $\mu$, where in each iteration (generation),
$\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-14]{genetic_algos}.
\\
\textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
performance metric the model resulting from the individual's HP configuration is evaluated on.
\\
\textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
index $k$ on are swapped, yielding two new ``child'' individuals.
\\
\textbf{Mutation} randomly changes some HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
on binary values.
\\
At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
\citep[p. 188]{simon2013evolutionary}

@@

Another popular EA approach is Differential Evolution. Its populaton is randomly initialized so the entire search space is covered.
In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
$\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also differ from $i$ and $F\in[0,2]$ is constant.
For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})^T$ with
\begin{equation}
  \upsilon_{i,g+1}^{(j)} = \begin{dcases}
    \nu_{i,g+1}^{(j)} & \text{if } u\le\text{CR}\text{ or } j=R \\
    \lambda_{i,g}^{(j)} & \text{else}
  \end{dcases}
\end{equation}
for a random index $R\in\{1,2,...,N\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}

@@

The kernel function can be freely chosen, a common choice is the Gaussian (Eq. \ref{eq-kernel-gaussian}) kernel with

@@

A drawback of Gaussian Processes is their poor scalability in number of data points and number of hyperparameters, although there exist workarounds approximating the full GP

@@

\subsection{Neural Architecture Search}
A special case of HPO is Neural Architecture Search (NAS). Because of the flexibility of neural network architectures (number of hidden layers, strength of dropout,
each layer can have a different numer of neurons, each layer can have a different activation function, etc.), traditional HPO methods cannot efficiently explore the
entire space.
In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and
image domains \citep{baker2017designingneuralnetworkarchitectures},  \citep{10.5555/3504035.3504375},  \citep{10.1145/3071178.3071229},
where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spaces}.
These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For feedforwad neural networks,
this could be a linear layer of fixed size and with a specific activation function,
for CNNs this could be a specific convolutional operation or pooling function, for instance.
Instead of tuning every hyperparameter, the sequential placement of the cells is then optimzed during NAS, e.g. via random search or Bayesian
Optimization. \citep[p. 67]{elsken_neural_2019}
\\
Treating a neural network as a sequence of cells instead of one fixed entity further allows exploiting this structure with special sequential optimization techniques.
\citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} and \citet[pp. 2-4]{Zoph_2018_CVPR} propose a cell search method using RNNs and reinforcement learning.
The cells in their approach are not fixed, but have hyperparameters themselves, such as the filter size and the stride size of a convolutional layer.
These hyperparameters are then predicted sequentially, given all the already predicted hyperparameters of previous layers. The RNN is trained using reinforcement
learning, where the actions are the different values the network can predict for a given hyperparameter and the reward function is the RNN's performance on held-out
validation data.
\\
The second approach to NAS is the \textbf{one-shot model}, which trains a ``fabric'', which can be understood as a DAG comprising all architectures of a given
search space. Each path through the DAG is one specific network architecture, where the nodes represent a layer of neurons and the edges in-between are operations
on the layers' values. The entire DAG has one designated input and one output node.
It is then trained just like a regular NN, after which the optimal path, i.e. network architecture, is selected from the fabric.
This has the advantage of, albeit being slower than training a single model, being far more efficient and less expensive than training all architectures encoded
in the fabric. \citep[pp. 1-2, p. 8]{saxena2017convolutionalneuralfabrics}
% don't necessarily talk about co-adaptation problem

@@

\subsubsection{A-posteriori}
\label{sec-moo-post}
The disadvantage of a-priori methods is the missing knowledge of the interplay between a hyperparameter configuration and the trained model's
performance across objectives:
we can either restrict the search space to enforce some will, such as only using a maximum of 50\% of features, optimize for performance and use the
resulting model, or leave the search space unrestricted, adjust the loss function to incorporate all objectives and take its optimum.
In neither case do we know the impact our a-priori trade-off has on the final performance.
In practical applications, however, the impact on performance across all objectives is what matters.
This is the main advantage of a-posteriori methods: instead of implicitly defining a trade-off prior to training, we still evaluate multiple HP configurations
but keep the set of non-dominated solutions instead of just one that satisfies the trade-off criterion.
This is beneficial as it makes the relationship visible, a practicioner can now see, for instance, that a slight decrease in one objective might translate to
a significant improvement in another that would have been missed if the problem was optimized with a-priori methods.
\\
Aside from the usual baselines grid and random search there are multi-objetive Bayesian Optimization adaptations, mainly using one of two approaches:
\begin{enumerate}[label*=\arabic*.]
  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augmented Tchebycheff function as
        scalarization to ensure the Pareto front is explored sufficiently.
  \item Fitting one surrogate per objective, then either 
  \begin{enumerate}[label*=\arabic*.]
    \item using one acquisition function per surrogate to return a set of promising candidate query points or
    \item using one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises
          the expected improvement of hypervolume.
  \end{enumerate}
\end{enumerate}

Lastly, there is also a multitude of multi-objective evolutionary algorithms (MOEA) to explore the Pareto front.
A key challenge for MOEAs is determining the fitness of an individual, as there are now multiple objectives that need to be incorporated instead of
one as in single-objective EA (SOEA).
We will outline the popular NSGA-II (Nondominated Sorting Genetic Algorithm II), which improves upon its predecessor NSGA \citep{NSGA} by making it
parameterless, ensuring elitism, and reducing the computational complexity of ranking of individuals \citep[p. 182]{NSGA_II}.
NSGA-II uses all the regular operators reproduction, mutation, and crossover that are already known from SOEA.
The difference is in the fitness function, which for NSGA-II comprises two parts: non-dominated sorting and crowding distance.
\\
\textit{Non-dominated sorting}, as described by \citet[p. 201]{genetic_algos} and \citet[pp. 183f]{NSGA_II} is an iterative procedure to ensure
elitist selection through ranking individuals by their fronts.
First, it assigns all individuals of the Pareto front a rank of 0. It then temporarily removes these individuals from the population and repeats
this procedure with an incremented rank number until no individuals are left.
\\
The second part, \textit{crowding distance} ensures sufficient diversity, i.e. exploration of the Pareto front.
The algorithm assigns each individual a score depending on how ``crowded'' the area in objective space around it is. Crowding distance is computed
as the mean side length of the cuboid spanned by its nearest neighbours as vertices in the objective space, where individuals without two neighbours
along some dimension are assigned an infinitely high distance value. The less the crowding distance, the more an area is crowded with other individuals. \citep[p. 185]{NSGA_II}
\\
NSGA-II then ranks individuals by their non-dominated sorting front ranks in ascending order and by their crowding distance in descending order as tie breaker.
This ranking is then used for selecting the $\mu$ best individuals to keep for the next generation.
Reproduction follows a binary tournament selection, where two individuals are randomly drawn from the population and the better one according to the ranking gets
included in the reproduction pool for mutation and crossover.
\\
Conceptually, the reasoning behind using non-dominated sorting to rank individuals is intuitive: the closer an individual's front is to the Pareto front, i.e.
the smaller its front rank, the closer it is to the true Pareto front.
The usage of crowding distance, on the other hand, is not as straightforward. To understand it, we recall the goal of MOO, which is not merely optimizing
hypervolume, but to approximate the true Pareto front well.
Considering the scenario of only having individuals of one area in the objective space included in the Pareto front estimate makes it very unstable: removing this
area (i.e. the solutions from this area) would collapse the entire front estimate.
Considering now the opposite, having multiple areas making up the Pareto front estimate, makes it stable: removing one area would not impact the estimate a lot,
as it's still held up by other individuals \citep[p. 185, pp. 189-192]{genetic_algos}.
A possible proxy for exploration of the Pareto front is the hypervolume contribution. The hypervolume contribution of individual
$\boldsymbol\lambda\in\hat{\mathcal{P}}$ is the difference between the hypervolume of the entire estimated Pareto front vs the hypervolume of the front without the
individual: $\text{CON}_{\hat{\mathcal{P}}}(\boldsymbol\lambda)=\text{HYP}(\hat{\mathcal{P}})-\text{HYP}(\hat{\mathcal{P}}\setminus\{\boldsymbol\lambda\})$,
where $\text{HYP}(\hat{\mathcal{P}})$ denotes the hypervolume induced by the Pareto set $\hat{\mathcal{P}}$ \citep[p. 384]{10.5555/1943267.1943271}.
